Extended Data

### 2021-11-15 doc/manuscript/Dow_manuscript_Nature.Rmd
finish reviewing highlights in google doc

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE [@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have a negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
Rather than lengthening the growing season, the primary effect of warming temperature on annual growth is a negative correlation with $T_{max}$ and associated drought stress during the peak growing season[@dorangeville_drought_2018], offsetting any benefit of an earlier onset of growth in the spring[@zhang_drought_2021].
Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis.
The effect of warm springs specifically in years with higher levels of precipitation during the peak growing season is not elucidated from this study, and should be examined in the future to clarify this relationship.

Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season (Fig. `r tree_rings`) than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are leading to a lengthening of the period from peak stem growth to the cessation of ecosystem C uptake.
As the extra C sequestration in years with warm springs does not substantially augment woody growth, it remains unclear how it is allocated within the ecosystem.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

@@

Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[**REFS**] are likely to alter the phenology of forest C cycling.
If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.

>>>>

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE [@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion, thus having negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
Rather than lengthening the growing season, the primary effect of warming temperature on annual growth is a negative correlation with $T_{max}$ and associated drought stress during the peak growing season[@dorangeville_drought_2018], offsetting any benefit of an earlier onset of growth in the spring[@zhang_drought_2021].
Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010] were uncorrelated within our dendrometer band analysis.
The effect of warm springs specifically in years with higher levels of precipitation during the peak growing season is not elucidated from this study, and should be examined in the future to clarify this relationship.

Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season (Fig. `r tree_rings`) than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect with findings that NEE increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of ecosystem C uptake.
As the extra C sequestration in years with warm springs does not substantially augment woody growth, it remains unclear how it is allocated within the ecosystem.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

@@

Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and carbon cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the phenology of forest C cycling.
If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and NEE [@churkina_spatial_2005].
Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.
